### 2015-10-07 paper-1/main.tex
Changes made based on Dan's and Michelle's suggestions.

\noindent${}^*$Email address for author: {\tt
michael.kane@yale.edu}.  This research was supported by grants 
	R01CA131301, R01CA157749, R01CA148996, R01CA168733 awarded by the National Cancer Institute, and support from the Yale Comprehensive Cancer Center.

@@

In the Simon 2-stage Phase II clinical trial [\ref{Simon 1989}], a small 
number of patients are enrolled onto a therapeutic 
protocol for an initial stage. For each patient, we assess either success 
or failure according to a 
clinically useful criteria. Success might include tumor response to treatment
or perhaps that the response is maintained for a minimum specified duration.
In addition to failing to achieve the primary endpoint, treatment failure 
might include adverse reaction to the drug or withdrawal due to toxicity. We 
use a large number of these early successes as evidence for whether to treat
additional patients in a second stage. A statistically small number of 
successes in Stage I would lead us to terminate the trial without entering 
Stage II.

Consider the Phase II clinical trial that investigated the efficacy of 
iniparib in patients with breast cancer gene-associated (BRCA) ovarian 
cancer [\ref{Sanofi 2013}]. In the first stage 12 patients were treated with 
iniparib (BSI-201/SAR240550). The endpoint was defined by Response Evaluation 
Criteria in Solid Tumor (RECIST). If at least two Stage I patients succeeded,
then the trial would have proceed to the next stage where additional patients 
were treated. If fewer than two of the 12 Stage I patients is successful then 
the trial would have terminated.

Our goal is to describe the distribution of the sample size of the design for 
planning purposes and to use this distribution to estimate the
success probability.  If all 12 patients are enrolled at once, 
as in the classic 
design, then the sample size is 12. However, in most clinical trials, the 
patients are enrolled sequentially, often with one patient's outcome realized 
before the next one enters the trial. In the present example, observing two 
successful patients allows us to proceed to the next stage. Similarly 11 
observed treatment failures also ends the stage. This sampling mechanism, in 
which the experiment ends as soon as any of the endpoints is reached, is 
call {\em curtailed sampling}. Under curtailed sampling the range of the 
sample size is between two and 12.

We assume each of the patient outcomes can be modeled as independent, 
identically distributed Bernoulli($p$) random variables. The realization of 
the trial is a sequence of these random variables that stops when either a 
specified number of success or failures is reached. As an example, suppose in 
the previously described trial, two successes were reached after enrolling 10 
patients (one in the third step and one at the $10^{th}$). The trial is 
visualized in Fig \ref{fig:kane_viz}. The vertical axis denotes the number of 
successful outcomes. The horizontal axis denotes the number of patients that 
have been enrolled. The horizontal and vertical line segments represent 
endpoints for the trial.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically 
distributed, Bernoulli random variables with $\mathbb{P}\{b_i=1\}=p$, for 
probability parameter $0\leq p \leq 1$. In the clinical trial setting 
$\,b_i = 0$ corresponds to an unsuccessful outcome.  Let $s$ and $t$ be 
positive integers.  Define the stopped negative binomial (SNB) random 
variable $Y$ as the smallest 
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either} 
$\,s\,$ successes {\em or} $\,t\,$ failures. The distribution of $\,Y\,$ has 
support on integer values in the range 
\begin{equation*}                                     %   (3)
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{P}[Y=k] = S(k, p, s)  [s \leq k] + T(k, p, t) [t \leq k ] 
\end{equation}
where 
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability of $s$ successful outcomes and
\begin{equation} \label{eqn:R}
T(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is the negative binomial probability of $t$ unsuccessful outcomes. 
From (\ref{eqn:pmf}) one can see that the mass at any point on the support 
of the distribution may come from either $X_k = s$, captured in the 
function $S$, or $X_k = k-t$, captured in the function $T$. The support of 
$S$ is defined by $s \leq S \leq s+t-1$ and the support of $T$ is defined 
by $t \leq T \leq s+t-1$.

@@

To show the result in Equation (\ref{eqn:pmf}) consider the 
process $\mathbf{X} = \{X(k) : k \in \mathbb{Z}_{\ge 0} \}$ with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \left[ k-t < X_k < s \right].
\end{equation*}
As before the $b_{k+1}$'s are distributed as Bernoulli($p$) for all $k$ in the 
domain of $\mathbf{X}$. The process can be conceptualized as a series of coin 
flips that stops when either $s$ heads or $t$ tails are reached. At each step 
a coin is flipped. If it is heads, the process advances one diagonally in the 
positive horizontal and vertical direction. Otherwise, it advances in the 
positive horizontal direction only. When the process stops, $X_k$'s and 
remains either $X_k = s$ or $X_k = k-t$. One example is given in 
Fig. \ref{fig:kane_viz}.

@@

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} I_{qe^x} (s, t) + \frac{qe^x}{1-pe^x} I_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MFG of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and this can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s I_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one 
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

Equation \ref{eqn:posterior} suggests an alternative construction for 
Phase II clinical trials.  Patients could be 
enrolled sequentially until there are enough patients to guarantee the 
desired level of power or until a specified number of adverse events are 
recorded. In cases such as these the endpoints $s$ and $t$ are known and 
new outcomes can be incorporated into the estimate of
$p$ to provide updated odds of successful enrollment. Later it will be shown 
that estimates of the distribution of $p$ can even be used after an endpoint 
has been reached to determine the amount of uncertainty in the endpoint.

>>>>

\noindent${}^*$Email address for author: {\tt
michael.kane@yale.edu}.  
This research was supported by grants R01CA131301, R01CA157749, R01CA148996, 
R01CA168733, and PC50CA196530 awarded by the National Cancer Institute, and 
support from the Yale Comprehensive Cancer Center.

@@

%In the Simon 2-stage Phase II clinical trial [\ref{Simon 1989}], a small 
%number of patients are enrolled onto a therapeutic 
%protocol for an initial stage. For each patient, we assess either success 
%or failure according to a 
%clinically useful criteria. Success might include tumor response to treatment
%or perhaps that the response is maintained for a minimum specified duration.
%In addition to failing to achieve the primary endpoint, treatment failure 
%might include adverse reaction to the drug or withdrawal due to toxicity. We 
%use a large number of these early successes as evidence for whether to treat
%additional patients in a second stage. A statistically small number of 
%successes in Stage I would lead us to terminate the trial without entering 
%Stage II.

Consider the design for a Phase II clinical trial to investigated the 
efficacy of iniparib in patients with breast cancer gene-associated (BRCA) 
ovarian cancer [\ref{Sanofi 2013}]. In the first stage, 12 patients were 
treated with iniparib (BSI-201/SAR240550). The endpoint was defined by 
Response Evaluation Criteria in Solid Tumor (RECIST). If at least two of
these patients respond favorably, then the trial will proceed to the next 
stage where additional patients are able to be treated. If fewer than two 
respond then the trial would be terminated.

The maxiumu sample size is 12 but the number of patients necessary to 
reach and endpoint can be less. Our goal is to describe the distribution of 
the enrollment size of the design for 
planning purposes and to use this distribution to estimate the
success probability.  If all 12 patients are enrolled at once, 
as in the classic 
design, then the sample size is 12. However, in most clinical trials, the 
patients are enrolled sequentially, often with one patient's outcome realized 
before the next one enters the trial. In the present example, observing two 
successful patients allows us to proceed to the next stage. Similarly 11 
observed treatment failures also ends the stage. This sampling mechanism, in 
which the experiment ends as soon as any of the endpoints is reached, is 
call {\em curtailed sampling}. Under curtailed sampling the range of the 
sample size is between two and 12.

We assume each of the patient outcomes can be modeled as independent, 
identically distributed Bernoulli($p$) random variables. The realization of 
the trial is a sequence of these random variables that stops when either a 
specified number of success or failures is reached. As in the 
previous example, two successes were reached after enrolling 10 
patients (one in the third step and one at the $10^{th}$). The trial is 
visualized in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of 
successful outcomes. The horizontal axis denotes the number of patients that 
have been enrolled. The horizontal and vertical line segments represent 
boundayr endpoints for the trial.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically 
distributed, Bernoulli random variables with $\mathbb{P}\{b_i=1\}=p$, for 
probability parameter $0\leq p \leq 1$. In the clinical trial setting 
$\,b_i = 0$ corresponds to an unsuccessful outcome.  Let $s$ and $t$ be 
positive integers.  Define the stopped negative binomial (SNB) random 
variable $Y$ as the smallest 
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either} 
$\,s\,$ successes {\em or} $\,t\,$ failures. The distribution of $\,Y\,$ has 
support on integer values in the range 
\begin{equation*}                                     %   (3)
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{P}[Y=k] = S(k, p, s)\left[s \leq k\right] + 
  S(k, 1-p, t) \left[ t \leq k \right]
\end{equation}
where 
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability truncated to a subset of the support 
- in this case, $s$ and $t$ respectively.
From Eqn.~\ref{eqn:pmf} one can see that the mass at 
any point on the support of the distribution may come from either 
$X_k = s$, captured in the function $S$, or $X_k = k-t$, captured in the 
function $T$. The support of 
$S$ is defined by $s \leq S \leq s+t-1$ and the support of $T$ is defined 
by $t \leq T \leq s+t-1$.

@@

To show the result in Eqn.~\ref{eqn:pmf} consider the 
process $\mathbf{X} = \{X(k) : k \in \mathbb{Z}_{\ge 0} \}$ with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \left[ k-t < X_k < s \right].
\end{equation*}
As before the $b_{k+1}$'s are distributed as Bernoulli($p$) for all $k$ in the 
domain of $\mathbf{X}$. The process can be conceptualized as a series of coin 
flips that stops when either $s$ heads or $t$ tails are reached. At each step 
a coin is flipped. If it is heads, the process advances one diagonally in the 
positive horizontal and vertical direction. Otherwise, it advances in the 
positive horizontal direction only. When the process stops
either $X_k = s$ or $X_k = k-t$. One example is given in 
Fig.~\ref{fig:kane_viz}.

@@

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} I_{qe^x} (s, t) + \frac{qe^x}{1-pe^x} I_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and this can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s I_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one 
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

Eqn.~\ref{eqn:posterior} suggests an alternative construction for 
Phase II clinical trials.  Patients could be 
enrolled sequentially until there are enough patients to guarantee the 
desired level of power or until a specified number of adverse events are 
recorded. In cases such as these the endpoints $s$ and $t$ are known and 
new outcomes can be incorporated into the estimate of
$p$ to provide updated odds of successful enrollment. Later it will be shown 
that estimates of the distribution of $p$ can even be used after an endpoint 
has been reached to determine the amount of uncertainty in the endpoint.
